SNP Analysis

### 2018-07-06 SNP_Prediction_Analysis.Rmd
processing per endo instead of per sample

@@

ptdirs = list.dirs('/Volumes/fh/fast/reid_b/collab/Killcoyne/Data/PerPatient', full.names=T, recursive=F)
patients = basename(ptdirs)

@@

#load('/Volumes/fh/fast/reid_b/collab/Killcoyne/SNP_R/allpts_ascat.Rdata', verbose=T)
load('~/Data/Ellie/Analysis/SNP/allpts_ascat.Rdata', verbose=T)
#patient.info = as.data.frame(read_xlsx('/Volumes/fh/fast/reid_b/collab/Killcoyne/SNP_Project/metadata_T1T2.xlsx'))
patient.info = as.data.frame(read_xlsx('~/Data/Ellie/Analysis/SNP/metadata_T1T2.xlsx'))
patient.info$UniqueSampleID = paste(patient.info$PatientID, patient.info$`Timepoint Code`, sep='_')
patient.info$Path.Status = patient.info$Status
patient.info[patient.info$PatientID %in% subset(patient.info, Pathology %in% c('IMC','HGD'), select='PatientID')[,1], 'Path.Status'] = 'P'

sample.list = read_xlsx('/Volumes/fh/fast/reid_b/collab/Killcoyne/Data/20180604_Reid_1M_SampleList.xlsx', sheet = 2)
nrow(sample.list)
colnames(sample.list)[3] = 'Total.SCA'

@@

In the sWGS data progressor samples show a variance of `r round(pv, 3)` and non-prog variance is `r round(npv,3)`

# Quick QC

@@

lowsca = subset(qcdata, Status == 'P' & SampleType == 'BE' & SCA.Ratio <= 0.05 & Ploidy <= 2)
write.table(lowsca, sep='\t', quote=F, row.names = F, file = '~/Data/Ellie/Analysis/SNP/low_sca_samples.txt')

@@

pt = sub('_.*', '', lowPloidyCN$Samplename[1])
pfiles = list.files(paste('/Volumes/fh/fast/reid_b/collab/Killcoyne/Data/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=T)
head(ascat.output$segments_raw)
seg = subset(ascat.output$segments_raw, grepl(lowPloidyCN$Samplename[1], sample))
seg$chr = factor(seg$chr, levels=c(1:22,'X','Y'))

plot.seg<-function(seg, incAB=F) {
  p = ggplot(seg) + facet_grid(~chr, space='free_x', scales='free_x') + 
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), color='darkgreen', size=3)

  if (incAB)
    p = p + geom_segment(aes(x=startpos, xend=endpos, y=nAraw+nBraw, yend=nAraw+nBraw), color='blue', size=3)

  p + theme(axis.text.x=element_blank() )
}

plot.seg(seg,incAB = T) + labs(title=paste('Low SCA Prog sample',lowPloidyCN$Samplename[1]), x='')

pt = sub('_.*', '', highPloidyCN$Samplename[1])
pfiles = list.files(paste('/Volumes/fh/fast/reid_b/collab/Killcoyne/Data/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=T)
head(ascat.output$segments_raw)
seg = subset(ascat.output$segments_raw, grepl(highPloidyCN$Samplename[1], sample))
seg$chr = factor(seg$chr, levels=c(1:22,'X','Y'))

plot.seg(seg, incAB = T) + labs(title=paste('High SCA Non-Prog sample',highPloidyCN$Samplename[1]), x='')

@@

These are the CN adjusted, winsorized LogR values tiled in 5Mb bins.

```{r, warning=F}

if (file.exists('tmp_seg_pt.Rdata')) {
  load('tmp_seg_pt.Rdata', verbose=T) 
} else {

@@

save(mergedSegs, mergedArms, file='tmp_seg_pt.Rdata')
}

rownames(mergedSegs) = sub('\\.LogR','', rownames(mergedSegs))
rownames(mergedArms) = sub('\\.LogR','', rownames(mergedArms))

# impute

# mergedSegs[is.na(mergedSegs)]
# mergedArms[is.na(mergedArms)]

# for (i in 1:nrow(mergedSegs)) 
#   mergedSegs[i,is.na(mergedSegs[i,])] = 0# mean(mergedSegs[i,],na.rm=T)
# 
# for (i in 1:nrow(mergedArms)) 
#   mergedArms[i,is.na(mergedArms[i,])] = 0 #mean(mergedArms[i,],na.rm=T)

@@

```{r normals, fig.width=6, fig.height=8, eval=F}
nm = melt(mergedSegs[grep('BLD|gastric', rownames(mergedSegs), ignore.case=T),])
be = melt(mergedSegs[grep('BLD|gastric', rownames(mergedSegs), ignore.case=T, invert=T),])

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric') + theme(axis.text.x = element_blank()),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples') + theme(axis.text.x = element_blank())
)

@@

Using a random normal patient and a known BE progresser patient

```{r rnormals, echo=F, fig.width=8, fig.height=8, eval=F}
randNormal = sample(grep('BLD', rownames(mergedSegs), ignore.case=T, value=T), 1)
pt = unlist(strsplit(randNormal, '_'))[1]
segments.list[[pt]] = subset(segments.list[[pt]], chr %in% c(1:22))
segments.list[[pt]]$chr = factor(segments.list[[pt]]$chr, levels=c(1:22), ordered = T)

grid.arrange(
  plot.seg(subset(segments.list[[pt]], sample == paste(randNormal,'.LogR',sep=''))) + labs(title='Median LRR, ASCAT segments', x='', subtitle=pt),
  ggplot(melt(as.matrix(mergedSegs[randNormal,])), aes(Var1, value)) + geom_point() + labs(title='Winsorized') + ylim(-1,1) + theme(axis.text.x=element_blank()),
top='Random BLD or gastric normal', bottom=randNormal)

randBE = '512_18762_133R_27.LogR'
#randBE = paste(sample(grep(paste('^',sample(subset(patient.info, Status == 'P')$PatientID, 1),sep=''), rownames(arrayDf), value=T), 1),'.LogR',sep='')
pt = unlist(strsplit(randBE, '_'))[1]
segments.list[[pt]] = subset(segments.list[[pt]], chr %in% c(1:22))
segments.list[[pt]]$chr = factor(segments.list[[pt]]$chr, levels=c(1:22), ordered = T)

grid.arrange(
  plot.seg(subset(segments.list[[pt]], sample == randBE)) +
    labs(title='Median LRR, ASCAT segments', subtitle=pt), 
  ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() + ylim(-1,1) + 
    theme(axis.text.x=element_blank()) + labs(title='Winsorized LRR, binned'),  
top='BE', bottom=randBE)

@@

I adjust the variance of each genomic region by the mean and sd extracted when I trained on the sWGS data. This seems to be a more fair comparison than scaling all the new data to itself, but the predictions don't really change.
```{r, warning=F}
# uv 
# ms  = t(apply(mergedSegs, 1, function(sample) {
#   sapply(1:ncol(mergedSegs), function(i){
#     unit.var(sample[i], z.mean[i], z.sd[i])
#   })
# }))
# 
# range(ms)
ms = mergedSegs
for (i in 1:ncol(ms)) {
  #ms[,i] = unit.var(ms[,i])
  ms[,i] = unit.var(ms[,i], z.mean[i], z.sd[i])
}
range(ms)
dim(ms)
# ma  = t(apply(mergedArms, 1, function(sample) {
#   sapply(1:ncol(mergedArms), function(i){
#     unit.var(sample[i], z.arms.mean[i], z.arms.sd[i])
#   })
# }))

ma = mergedArms
for (i in 1:ncol(ma)) {
  #ma[,i] = unit.var(ma[,i])
  ma[,i] = unit.var(ma[,i], z.arms.mean[i], z.arms.sd[i])
}
range(ma)
dim(ma)

nm = melt(ms[grep('BLD|gastric', rownames(ms), ignore.case=T),])
be = melt(ms[grep('BLD|gastric', rownames(ms), ignore.case=T, invert=T),])

grid.arrange(
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples'),
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(titls='Blood/gastric')
)

@@

ggplot(melt(arrayDf), aes(Var2, value, group=Var2)) + geom_point() + labs(title='All, scaled')

@@

offset = log(0.0225)
preds = cbind.data.frame(rownames(prob),prob, rr)
colnames(preds) = c('Samplename','Prob', 'RR')

# preds = preds %>% dplyr::mutate( 
#   'Adj. RR'=RR+offset,  
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

@@

@@

A quick look at the two random samples from earlier.  The random normal shows a very low probablity before and after adjustment:
`r pander(subset(preds, Samplename == randNormal))`

The random BE shows a very high probability:
`r pander(subset(preds, Samplename == sub('\\.LogR','',randBE)))`

```{r, eval=F}
hist(subset(preds, Samplename %in% lowsca$Samplename)$Prob)

x = subset(segments.list$`891`, sample == '891_18574_100O_33.LogR') 
x$chr = factor(x$chr, levels=c(1:22,'X','Y'), ordered = T)
p1 = ggplot(x) + facet_grid(~chr, space='free_x', scales='free_x') + 
  geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + labs(title='891_18574_100O_33 medLRR')

x = melt(ms['891_18574_100O_33',])
x$loc = rownames(x)
x = rbind(x,cbind.data.frame('loc'=paste(c(1:22), '0-1', sep=':'), 'value'=0))

locs = do.call(rbind.data.frame, strsplit(x$loc, ':|-'))
colnames(locs) = c('chr','start','end')
locs[] = lapply(locs[], function(y) as.numeric(as.character(y)))
x = cbind(x,locs)
x$chr = factor(x$chr, levels=c(1:22), ordered=T)
x = arrange(x,chr)

p2 = ggplot(x) + facet_grid(~chr, space='free_x', scales='free_x') + 
  geom_segment(aes(x=start, xend=end, y=value, yend=value), size=3) + labs(title='891_18574_100O_33 binned')

grid.arrange(p1,p2,nrow=2)
```

## Evaluating normals vs BE

```{r, warning=F, fig.height=8}
normals = subset(preds, grepl('BLD|gastric', Samplename, ignore.case=T))
grid.arrange(
  ggplot(normals, aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
    scale_fill_gradientn(colors = myPal,  name='') + 
    labs(title='Predictions on BLD/Gastric', y='n Samples', x='Unadjusted Probability'),
top='BLD/gastric normals' )

be = subset(preds, !grepl('BLD|gastric', Samplename, ignore.case=T))
grid.arrange(
  ggplot(be, aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
    scale_fill_gradientn(colors = myPal,  name='') + 
    labs(title='Predictions on BE', y='n Samples', x='Unadjusted Probability'),
top='BE samples' )
pred.qc = base::merge(preds, qcdata, by='Samplename')
```

So how do these high normals look?
```{r, eval=F}
preds
qcdata
nrow(preds)

rc = Hmisc::rcorr( as.matrix(pred.qc[,c(2,3,6,7,8,12)]) )

GGally::ggcorr( as.matrix(pred.qc[,c(2,6,7,8,12)]) , label=T, palette='RdBu')

length(which(normals$Prob < 0.5))/nrow(normals)

highNmSamples = subset(normals, Prob >= 0.5)
subset(qcdata, Samplename %in% highNmSamples$Samplename)
# Low variance overall
mean(apply(mergedSegs[highNmSamples$Samplename,], 1, var))

# Most of the gastric "normals" show up, though that drops to 3% when adjusted
length(grep('gastric', highNmSamples$Samplename, ignore.case=T))/length(grep('gastric', rownames(mergedSegs), ignore.case=T))
length(grep('gastric', subset(normals, `Adj. Prob` >= 0.5)$Samplename, ignore.case=T))/length(grep('gastric', rownames(mergedSegs), ignore.case=T))

# About 8% of the blood normals do as well, but drops to <1% when adjusted
length(grep('BLD', highNmSamples$Samplename, ignore.case=T))/length(grep('BLD', rownames(mergedSegs), ignore.case=T))
length(grep('BLD', subset(normals, `Adj. Prob` >= 0.5)$Samplename, ignore.case=T))/length(grep('BLD', rownames(mergedSegs), ignore.case=T))
```

Compare BLD to BE
```{r, echo=F, fig.height=8}

unadj = length(unique(sapply(rownames(be[which(be$Prob > 0.5),]), function(x) unlist(strsplit(x, '_'))[1])))/length(patients)
#adj = length(unique(sapply(rownames(be[which(be$`Adj. Prob` > 0.5),]), function(x) unlist(strsplit(x, '_'))[1])))/length(patients)
#Adjusted, `r round(adj, 2)*100`% have a >50% probability.
```

Unajdusted `r round(unadj,2)*100`% of BE patients have a >50% probability of progression.

@@

```{r roc, echo=T}
#preds.info = base::merge(sample.info, preds, by='Samplename')
rows = grep('BLD|gastric', pred.qc$Samplename, ignore.case=T)
normals = pred.qc[rows,]
pred.qc = pred.qc[-rows,]
nrow(pred.qc)

# Still missing some SNP data  FIX THIS CHECK
si = do.call(rbind.data.frame,c(strsplit(as.character(pred.qc$Samplename), '_'), stringsAsFactors=F) )
colnames(si) = colnames(sample.info)[1:4]

pred.qc = cbind(si, pred.qc)

setdiff(unique(patient.info$PatientID),unique(pred.qc$PatientID))

pred.qc$Status = factor(pred.qc$Status)

roc = pROC::roc(Status ~ Prob, data=pred.qc, auc=T, ci=T, of='thresholds')
roc.plot(roc)

rocNoLow = pROC::roc(Status ~ Prob, data=subset(pred.qc, !Samplename %in% c(lowsca$Samplename,highPloidyCN$Samplename)), auc=T, ci=T, of='thresholds')
roc.plot(rocNoLow)

head(pred.qc)

per_pt = pred.qc %>% group_by(PatientID, Timepoint, Status) %>% dplyr::summarise(
  max.Prob = max(Prob), max.RR = max(RR)
)

rocpp = pROC::roc(Status ~ max.Prob, data=per_pt, auc=T, ci=T, of='thresholds')
roc.plot(rocpp)

per_pt_ex = subset(pred.qc, !Samplename %in% c(lowsca$Samplename,highPloidyCN$Samplename)) %>% group_by(PatientID, Timepoint, Status) %>% dplyr::summarise(
  max.Prob = max(Prob), max.RR = max(RR)
)

rocppex = pROC::roc(Status ~ max.Prob, data=per_pt_ex, auc=T, ci=T, of='thresholds')
roc.plot(rocppex)

roct2 = pROC::roc(Status ~ Prob, data=subset(pred.qc, Timepoint != 'T1' & !Samplename %in% c(lowsca$Samplename,highPloidyCN$Samplename)), auc=T, ci=T, of='thresholds')
roc.plot(roct2)

>>>>

plot.sample<-function(segs, mer)

@@

#ptdirs = list.dirs('/Volumes/fh/fast/reid_b/collab/Killcoyne/Data/PerPatient', full.names=T, recursive=F)
#patients = basename(ptdirs)

@@

load('~/Data/Reid_SNP/PerPatient/allpts_ascat.Rdata', verbose=T)

patient.info = as.data.frame(read_xlsx('~/Data/Ellie/Analysis/SNP/metadata_T1T2.xlsx'))
patient.info$UniqueSampleID = paste(patient.info$PatientID, patient.info$`Timepoint Code`, sep='_')
patient.info$Path.Status = patient.info$Status
patient.info[patient.info$PatientID %in% subset(patient.info, Pathology %in% c('IMC','HGD'), select='PatientID')[,1], 'Path.Status'] = 'P'

sample.list = read_xlsx('~/Data/Ellie/Analysis/SNP/20180604_Reid_1M_SampleList.xlsx', sheet = 2)
nrow(sample.list)
colnames(sample.list)[3] = 'Total.SCA'

@@

## Patient info

patient.info = patient.info %>% rowwise() %>% dplyr::mutate( Endoscopy = ifelse(Timepoint == 'T1', 1, 2) )

sm = patient.info %>% group_by(Status, Endoscopy, Pathology) %>% dplyr::tally() %>% group_by(Status, Endoscopy) %>% dplyr::mutate(freq=n/sum(n) )
sm$Pathology = factor(sm$Pathology, levels=c('NDBE','ID','LGD','HGD','IMC'), ordered = T)

ggplot(sm, aes(Pathology, freq, group=Endoscopy, fill=factor(Endoscopy))) + facet_grid(~Status) + geom_col(position = position_dodge()) + ylim(0,1) + title('Pathology frequency per timepoint')
```

There's a significant difference in the numbers of NDBE or HGD samples between the groups
```{r}
chisq.test(table(patient.info$Pathology, patient.info$Status)['HGD',])
chisq.test(table(patient.info$Pathology, patient.info$Status)['NDBE',])

# Most progressors are HGD at timepoint 1
with(subset(patient.info, Endoscopy == 1), table(Pathology, Status))['HGD',]/table(subset(patient.info, Endoscopy == 1)$Status)
```

# Quick QC 
In the sWGS data progressor samples show a variance of `r round(pv, 3)` and non-prog variance is `r round(npv,3)`

@@

lowsca = subset(qcdata, Status == 'P' & SampleType == 'BE' & SCA.Ratio <= 0.05 & Ploidy <= 2)

@@

pt = '403'
pfiles = list.files(paste('~/Data/Reid_SNP/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=T)

seg = ascat.output$segments_raw
seg$chr = factor(seg$chr, levels=c(1:22,'X','Y'))
ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free') + 
  geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), color='darkgreen', size=3) + 
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA', subtitle='patient 403')

pt = '512'
pfiles = list.files(paste('~/Data/Reid_SNP/PerPatient', pt, sep='/'), full.names = T)
load(grep('ascat.Rdata', pfiles, value=T), verbose=T)

seg = ascat.output$segments_raw
seg$chr = factor(seg$chr, levels=c(1:22,'X','Y'))
ggplot(seg) + facet_grid(sample~chr, space='free_x', scales='free') + 
  geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), color='darkgreen', size=3) + 
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples', subtitle='patient 512')

@@

CN Adjustment uses Annalise's whole-genome non-progressor samples to generate a set of mean and SD values per copy-number state. The median LogR values are adjusted using the mean and SD based on the CN state.  The adjusted LogR values are then winsorize per sample, and tiled in 5Mb bins.

```{r, warning=F, echo=F}

if (file.exists('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata')) {
  load('~/Data/Reid_SNP/PerPatient/tmp_seg_pt.Rdata', verbose=T) 
} else {

@@

save(mergedSegs, mergedArms, file='tmp_seg_pt.Rdata')
}
rownames(mergedSegs) = sub('\\.LogR','', rownames(mergedSegs))
rownames(mergedArms) = sub('\\.LogR','', rownames(mergedArms))

## Per endoscopy take the average or max bin value as an in silico pooling
sample.info$Level = toupper(sample.info$Level)

range(mergedSegs[1,])
range(x)
plot( x )

x = (mergedSegs[1,]-mean(apply(allDf[,1:ncol(mergedSegs)], 1, mean)))/mean(apply(allDf[,1:ncol(mergedSegs)], 1, sd))
  #sd(apply(allDf[,1:ncol(mergedSegs)], 1, sd))

# names = colnames(mergedSegs)
# mergedSegs = t(apply(mergedSegs, 1, scale, center=T))
# colnames(mergedSegs) = names
# 
# names = colnames(mergedArms)
# mergedArms = t(apply(mergedArms, 1, scale, center=T))
# colnames(mergedArms) = names

samples = unique(subset(sample.info, !Level %in% c('BLD','GASTRIC'), select=c("PatientID", "EndoID")))

meanBinSegs = as.data.frame(matrix(ncol=ncol(mergedSegs), nrow=0, dimnames=list(c(), colnames(mergedSegs))))
meanBinArms = as.data.frame(matrix(ncol=ncol(mergedArms), nrow=0, dimnames=list(c(), colnames(mergedArms))))

mean.qcdata = data.frame(matrix(ncol=8, nrow=0, dimnames=list(c(), c('Samplename','Status','SampleType','Timepoint','mean.sca','mean.ploidy','mean.fit','mean.purity'))))

for (i in 1:nrow(samples)) {
  samples[i,]
  print(paste(samples$PatientID[i], '(',i,')'))

  #rows = subset(sample.info, PatientID == samples$PatientID[i] & EndoID == samples$EndoID[i] )$Samplename
  ptendoID = paste(samples$PatientID[i], samples$EndoID[i], sep='_')

  mean.qcdata[i,] = 
    cbind('Samplename'=ptendoID, subset(qcdata, SampleType == 'BE' & grepl(paste(samples$PatientID[i], samples$EndoID[i], sep='.*'), Samplename)) %>% group_by(Status, SampleType, Timepoint) %>%
    dplyr::summarise(mean.sca=mean(SCA.Ratio), mean.ploidy=mean(Ploidy), mean.fit=mean(Goodnessoffit), mean.purity=mean(Purity)))

  rows = grep(paste(samples$PatientID[i], samples$EndoID[i], sep='.*'), rownames(mergedSegs), value=T)

  sharedR = intersect(rows, rownames(mergedSegs))
  if (length(rows) != length(sharedR)) {
      warning(paste("Missing samples for", endo$PatientID))
      rows = sharedR
  }

  #meanBinSegs[ptendoID,] = apply(mergedSegs[rows,,drop=F], 2, mean)
  meanBinSegs[ptendoID,] = apply(mergedSegs[rows,,drop=F], 2, function(x) x[which.max(abs(x))] )
  #meanBinArms[ptendoID,] = apply(mergedArms[rows,,drop=F], 2, mean)
  meanBinArms[ptendoID,] = apply(mergedArms[rows,,drop=F], 2, function(x) x[which.max(abs(x))] )
}

normalSegs = mergedSegs[grep('BLD|gastric',rownames(mergedSegs),ignore.case=T),]
normalArms = mergedArms[grep('BLD|gastric',rownames(mergedArms),ignore.case=T),]

meanBinSegs = as.matrix(meanBinSegs) 
meanBinArms = as.matrix(meanBinArms)

dim(meanBinSegs)

lowSeg = mergedSegs[grep('^403', rownames(mergedSegs), value=T),,drop=F]
p1 = ggplot(melt(lowSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA samples', subtitle='patient 403')
lowSeg = meanBinSegs[grep('^403', rownames(meanBinSegs), value=T),,drop=F]
p2 = ggplot(melt(lowSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Low SCA samples, mean bins', subtitle='patient 403')

grid.arrange(p1,p2)

highSeg = mergedSegs[grep('^512', rownames(mergedSegs), value=T),,drop=F]
p1 = ggplot(melt(highSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples', subtitle='patient 512')
highSeg = meanBinSegs[grep('^512', rownames(meanBinSegs), value=T),,drop=F]
p2 = ggplot(melt(highSeg), aes(Var2, value)) + facet_grid(~Var1, space='free_x', scales='free') + geom_point() +
  theme(axis.text.x=element_blank() ) + labs(title='Mixed SCA samples, mean bins', subtitle='patient 512')

grid.arrange(p1,p2)
```

@@

```{r normals, fig.width=6, fig.height=8, eval=F}
nm = melt(normalSegs)
be = melt(mergedSegs[grep('BLD|gastric', rownames(mergedSegs), ignore.case=T, invert=T),])
beM = melt(meanBinSegs[grep('BLD|gastric', rownames(meanBinSegs), ignore.case=T, invert=T),])

lims = range(nm$value, be$value, beM$value)

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric') + theme(axis.text.x = element_blank()) + ylim(lims),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples') + theme(axis.text.x = element_blank()) + ylim(lims),
  ggplot(beM, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples, mean per endo') + theme(axis.text.x = element_blank()) + ylim(lims)
)

@@

Using a random normal sample and a known BE progresser sampe

```{r rnormals, echo=F, fig.width=8, fig.height=8, eval=F}
randNormal = sample(rownames(normalSegs), 1)
pt = unlist(strsplit(randNormal, '_'))[1]
segs = subset(segments.list[[pt]], chr %in% c(1:22) & grepl(randNormal, sample))
segs$chr = factor(segs$chr, levels=c(1:22), ordered = T)

lims = range(segs$medLRR, mergedSegs[randNormal,])
grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free') + ylim(lims) +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='ASCAT median LRR', subtitle=pt),
  ggplot(melt(as.matrix(mergedSegs[randNormal,])), aes(Var1, value)) + geom_point() + ylim(lims) +
    labs(title='CN Adjusted & winsorized') + theme(axis.text.x=element_blank()),
top='Random BLD or gastric normal', bottom=randNormal)

randBE = '512_18762_133R_27.LogR'
pt = unlist(strsplit(randBE, '_'))[1]
segs = subset(segments.list[[pt]], chr %in% c(1:22) & grepl(randBE, sample))
segs$chr = factor(segs$chr, levels=c(1:22), ordered = T)

lims = range(segs$medLRR, mergedSegs[sub('\\.LogR','',randBE),])
grid.arrange(
  ggplot(segs) + facet_grid(~chr, space='free_x', scales='free') + ylim(lims) +
    geom_segment(aes(x=startpos, xend=endpos, y=medLRR, yend=medLRR), size=3) + 
    theme(axis.text.x=element_blank() ) + labs(title='Median LRR, ASCAT segments', subtitle=pt), 
  ggplot(melt(as.matrix(mergedSegs[sub('\\.LogR','',randBE),])), aes(Var1, value)) + geom_point() + ylim(lims) + 
    theme(axis.text.x=element_blank()) + labs(title='Winsorized LRR, binned'),  
top='BE', bottom=randBE)

@@

Adjust the variance of each genomic region by the mean and sd extracted when I trained on the sWGS data. This seems to be a more fair comparison than scaling all the new data to itself, but the predictions don't really change.

Using the mean bin values.
```{r, warning=F}
adjust.cols<-function(mtx, means=NULL, sds=NULL) {
  if (!is.null(means)) {
    if (ncol(mtx) != length(means))
      stop("Vector of means needs to match the columns in the matrix")
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i], means[i], sds[i])
  } else {
    for (i in 1:ncol(mtx)) 
      mtx[,i] = unit.var(mtx[,i])
  }
  return(mtx)
}

ns = adjust.cols(normalSegs, z.mean, z.sd)
na = adjust.cols(normalArms, z.arms.mean, z.arms.sd)

ms = adjust.cols(meanBinSegs, z.mean, z.sd)
ms = adjust.cols(mergedSegs, z.mean, z.sd)
range(ms)
dim(ms)

ma = adjust.cols(meanBinArms, z.arms.mean, z.arms.sd)
ma = adjust.cols(mergedArms, z.arms.mean, z.arms.sd)
range(ma)
dim(ma)

nm = melt(ns)
be = melt(ms[grep('BLD|gastric', rownames(ms), ignore.case=T, invert=T),])

grid.arrange(
  ggplot(nm, aes(Var2, value, group=Var2)) + geom_point() + labs(title='Blood/gastric') + 
    theme(axis.text.x=element_blank()) +ylim(range(nm$value, be$value)),
  ggplot(be, aes(Var2, value, group=Var2)) + geom_point() + labs(title='BE samples') + 
    theme(axis.text.x=element_blank()) + ylim(range(nm$value, be$value)), top='Unit scaled')

@@

cxN = score.cx(ns, 1)
arrayDfNorm = subtract.arms(ns, na)
arrayDfNorm = cbind(arrayDfNorm, 'cx'=unit.var(cxN, mn.cx, sd.cx))

probN = predict(fitV, newx=arrayDfNorm, s=l, type='response')
length(which(probN > 0.5))/nrow(probN) # FPR
length(which(probN <= 0.5))/nrow(probN)

@@

preds = cbind.data.frame(rownames(prob),prob, rr)
colnames(preds) = c('Samplename','Prob', 'RR')

#offset = log(0.0225)
# preds = preds %>% dplyr::mutate( 
#   'Adj. RR'=RR+offset,  
#   'Adj. Prob'=1/(1+exp(-RR+abs(offset)) )
# )

@@

#randNormal = paste(unlist(strsplit(randNormal, '_'))[c(1,3)], collapse='_')
#randBE = paste(unlist(strsplit(randBE, '_'))[c(1,3)], collapse='_')
```

A quick look at the two random samples from earlier.  
`r pander(subset(preds, Samplename == randNormal), caption='Random normal')`

The random BE shows a very high probability:
`r pander(subset(preds, Samplename == randBE), caption='Random BE')`

## Evaluating predictions

```{r, warning=F, fig.height=8}
# normals = subset(preds, grepl('BLD|gastric', Samplename, ignore.case=T))
# grid.arrange(
#   ggplot(normals, aes(Prob)) + geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
#     scale_fill_gradientn(colors = myPal,  name='') + 
#     labs(title='Predictions on BLD/Gastric', y='n Samples', x='Unadjusted Probability'),
# top='BLD/gastric normals' )
# 
# normals[which(normals$Prob > 0.5),]

pred.qc = base::merge(preds, mean.qcdata, by='Samplename')
#pred.qc = base::merge(preds, qcdata, by='Samplename')

prog = subset(pred.qc, !grepl('BLD|gastric', Samplename, ignore.case=T) & Status == 'P')
ggplot(prog, aes(Prob)) + facet_grid(~Timepoint) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on progressors', y='n Samples', x='Unadjusted Probability')

nonprog = subset(pred.qc, !grepl('BLD|gastric', Samplename, ignore.case=T) & Status == 'NP')
ggplot(nonprog, aes(Prob)) + facet_grid(~Timepoint) + 
  geom_histogram(aes(fill=..x..), breaks=seq(0,1,0.1) , show.legend = F) +
  scale_fill_gradientn(colors = myPal,  name='') + 
  labs(title='Predictions on non-progressors', y='n Samples', x='Unadjusted Probability')

@@

@@

```{r roc, echo=T}
si = do.call(rbind.data.frame,c(strsplit(as.character(pred.qc$Samplename), '_'), stringsAsFactors=F) )
if (ncol(si) == 4) {
  colnames(si) = colnames(sample.info)[1:4]
} else {
  colnames(si) = colnames(sample.info)[1:2]
}
pred.qc = cbind(si, pred.qc)
pred.qc = base::merge( pred.qc, patient.info[,c('PatientID','Pathology')], by='PatientID')

rows = grep('BLD|gastric', pred.qc$Samplename, ignore.case=T)
if (length(rows) > 0) {
  normals = pred.qc[rows,]
  pred.qc = pred.qc[-rows,]
}
nrow(pred.qc)

# Three groups, prog, non prog (max LGD), non prog (max HGD)
pred.qc = pred.qc %>% group_by(Status, PatientID) %>% dplyr::mutate( exclude=(Status == 'NP' & length(which(Pathology == 'HGD')) > 0) )
pred.qc$Status = factor(pred.qc$Status)

# Exclude those odd tetraploidy NPs
pred.qc[which(with(pred.qc, (Samplename %in% highPloidyCN$Samplename))), 'exclude'] = T
pred.qc[which(pred.qc$PatientID %in% subset(pred.qc, Samplename %in% highPloidyCN$Samplename)$PatientID), 'exclude'] = T

# Exclude HGD NP patients
pred = subset(pred.qc, !exclude)

roc = pROC::roc(Status ~ Prob, data=pred, auc=T, ci=T, of='thresholds')
roc.plot(roc)

#rocNoLow = pROC::roc(Status ~ Prob, data=subset(pred, !Samplename %in% c(lowsca$Samplename)), auc=T, ci=T, of='thresholds')
#roc.plot(rocNoLow)

per_pt = pred %>% group_by(PatientID, Timepoint, Status) %>% dplyr::summarise( max.Prob = max(Prob), max.RR = max(RR) )
rocpp = pROC::roc(Status ~ max.Prob, data=per_pt, auc=T, ci=T, of='thresholds')
roc.plot(rocpp)

per_pt_ex = subset(pred, !Samplename %in% c(lowsca$Samplename,highPloidyCN$Samplename)) %>% group_by(PatientID, Timepoint, Status) %>% dplyr::summarise(
  max.Prob = max(Prob), max.RR = max(RR)
)
rocppex = pROC::roc(Status ~ max.Prob, data=per_pt_ex, auc=T, ci=T, of='thresholds')
roc.plot(rocppex)

roct2 = pROC::roc(Status ~ Prob, data=subset(pred, Timepoint != 'T1' & !Samplename %in% c(lowsca$Samplename)), auc=T, ci=T, of='thresholds')
roc.plot(roct2)
